# One-step-ahead residual computation steps for mortals

For `"oneStepGeneric"`, which is probably the most instructive one to work
through, here are the steps to calculate the OSA residuals:

1. Map all fixed and random parameters to estimated values.

2. Add your data as a parameter vector(!).

3. Create an indicator vector 'keep' that can turn on or off a given data
   element in the likelihood by multiplying a 1 or 0 with the likelihood value.
   This will also be a parameter so we can easily pass it to our nll function
   calls.

4. Now we will iterate through the observations in order.

5. For each observation, we need the CDF of that observation conditional on
   previous data points with all parameters at their MLEs. There are many ways
   to get there. We'll continue with the oneStepGeneric algorithm including all
   tricks for staying in log space.

6. Build a new TMB object with the new map and new parameters (now including
   data as parameters and a 'keep' parameter vector).

7. Create a function that describes the negative log likelihood over a range of
   possible values for the given observation conditional on all the above. For
   `splineApprox = TRUE` this uses a spline on `tmbprofile(..., slice = TRUE)`
   to limit the number of nll function evaluations, otherwise this uses
   `Vectorize()` on the TMB nll `fn()` directly.

8. Now, we need to integrate that function up to the data point and to the
   right of that data point to get the CDF values for our data point we need to
   work with. For computational reasons, of course, TMB does this in log space.
   Here's how that happens.

In the following:

* `spline` is our function that returns the nll at a given value of our data
  point `x` (conditional as above)
* nll is the negative log likelihood at the observed value `x` (always
  conditional as above)
* `discrete` is `TRUE` (1) if discrete and `FALSE` (0) if continuous

```{r, eval=FALSE}
F1 <- integrate(function(x)exp(-(spline(x) - nll)),
                                spline.range[1], # where to start integration
                                obs[index] # where to end integration: the observation value
                                )$value
F2 <- integrate(function(x)exp(-(spline(x) - nll)),
                                obs[index] + discrete, # start at observation
                                spline.range[2] # to the right end
                                )$value
```

Then get back to a negative log CDF with:

```{r, eval=FALSE}
nlcdf.lower = nll - log(F1)
nlcdf.upper = nll - log(F2)
```

I *think* the above is basically what's described on the right side of equation
8 in Thygesen et al. as the way we can get at the CDFs with the already
existing model.

For `"cdf"` OSA type, TMB obtains the `nlcdf.lower` and `nlcdf.upper` directly from
the user-supplied .cpp template where these must be defined.

9. Do the above for each successive data point. As an example, for the 2nd data
   point, `keep` gets turned on (set to 1) for data points 1 and 2 and we do
   our integration of the nll across possible values of data point 2. For data
   point 3, `keep` gets turned on for data points 1, 2, and 3 and we do our
   integration of the nll across possible values of data point 3.

10. Once we've iterated over all observations that we're interested in, we can
   calculate the CDF of the current observation given past observations as:

```{r, eval=FALSE}
Fx <- 1 / ( 1 + exp(nlcdf.lower - nlcdf.upper))
```

FIXME: I *think* the `exp(nlcdf.lower - nlcdf.upper)` part is the $P^M$ ratio part of
equation 8 in Thygesen et al. calculated in log space but I'm still trying to
understand the `1 / 1 +` part. I can see why we need that to ensure we have
a variable between 0 and 1, i.e. the uniform distribution. Maybe this part is
clear to someone else?

`Fx` should be uniformly distributed if the model is consistent with the data.
This is $U_i$ in Thygesen et al. (2017).

I'm going to skip over the discrete observation randomization for simplicity,
but this part is basically the same as any randomized quantile approach.

11. Our residual is then the following as in any quantile residual approach:

```{r, eval=FALSE}
residual <- qnorm(Fx)
```

